Nonlinear methods to quantify Movement Variability in Human-Humanoid Interaction Activities

### 2021-03-18 docs/manuscript/main.tex
fixes references for figure02

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Experiment} \label{sec:experiment}
We conducted an experiment in the context of human-humanoid imitation (HHI) 
activities where participants were asked to imitate simple horizontal and 
vertical arm movements performed by NAO, a humanoid robot \cite{gouaillier2009}.
Such simple movements were repeated ten times for the participant 
who copied NAO's arm movements in a face-to-face imitation activity.
Also, wearable inertial measurement unit (IMU) sensors were attached 
to the right hand of the participant and to the left hand of the robot 
(Figure~\ref{fig:hri} A,C). Data were then collected with four NeMEMSi 
IMU sensors with sampling rate of 50Hz provinding tri-axial data of the 
accelerometer, gyroscope and magnetometer sensors and quaternions 
\cite{Comotti2014}.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
  \centering
\includegraphics[width=1.0\textwidth]{figures/fig02/versions/drawing-v00.png}
    \caption{
	{\bf Human-humanoid imitation activities.} 
		Face-to-face human-humanoid imitation (HHI) activities for (A.1) HHI of horizontal arm movement, (B.1) Humanoid horizontal arm movement, (C.1) HHI of vertical arm movement, and (D.1) Humanoid vertical arm movement.
	{\bf Time series duration for arm movements.}
    	Time series of smoothed data from gyroscope sensor for different speed arm movements performed by NAO: (A.2) Horizontal Normal arm movement (HN), (B.2) Horizontal Faster arm movement (HF), (C.2) Vertical Normal arm movement (VN) and (D.2) Vertical Faster arm movement (VF).
		(A.2) shows window sizes for 2-seconds (100 samples), 5-seconds (250 samples), 10-seconds (500 samples) and 15-seconds (750 samples) which are also presented in (B.2), (C.2) and (D.2).
	{\bf Time series for horizontal and vertical arm movements.}
		(A.3/D.3) raw-normalised (sg0zmuv), (B.3/E.3) normalised-smoothed 1 (sg1zmuv) and (C.3/F.3) normalised-smoothed 2 (sg2zmuv).
		Time series are only for three participants	($p01$, $p02$, and $p03$) for horizontal and vertical arm movements in normal and faster velocity (HN, HF, VN, VF) with the normalised GyroZ or GyroY axis and with one sensor attached to the participant (HS01) and other sensor attached to the robot (RS01).
	Code and data to reproduce the figure is available in \cite{srep2021}.
}
    \label{fig:hri}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

@@

\subsection*{Human-humanoid imitation activities}
For human-humanoid imitation (HHI) activities four neMEMSi sensors were used,
two of which were attached to the right hand of the participant and the 
other two to the left hand of the humanoid robot.
Then, each participant was asked to imitate repetitions of simple horizontal
and vertical arm movements performed by the humanoid robot in the following 
conditions:
(i) ten repetitions of horizontal arm movement at normal (HN) 
	and faster (HF) speed (Figure~\ref{fig:hri} A), and
(ii) ten repetitions of vertical arm movement at normal (VN) 
	and faster (VF) speed (Figure~\ref{fig:hri} C).
The normal and faster speed of arm movements is defined by the duration 
in number of samples of one repetition of NAO's arm movements. 
We select NAO's arm movements duration to distinguish between normal and 
faster arm movements as NAO's movements have less variation 
for such repetitive movements. 
The duration for one repetition of the horizontal 
arm movement at normal speed, HN, is 5 seconds considering 
that each repetition last around 250 samples.
For horizontal arm movement at faster speed, HF, each repetition were performed 
in 2 seconds which correspond to 90 samples of data.
The vertical arm movement at normal speed, VN, were performed  in 6 seconds 
which is around 300 samples of data. For vertical arm movement at 
faster speed, VF, each repetition lasts about 2.4 seconds which correspond 
to 120 samples of data.
To visualise the distinction between normal and faster speed for horizontal 
and vertical arm movements, Fig~\ref{fig:sts} shows smoothed time series 
for axes Z and Y of the gyroscope sensors with four window lengths: 
2-sec (100-samples), 5-sec (250-samples), 10-sec (500-samples) 
and 15-sec (750-samples).
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/experiment/sts/pdf/sts}
%    \caption{
%	{\bf Time series duration of horizontal and vertical arm movements.}
%    		Time series of smoothed data from gyroscope sensor
%		for different speed arm movements performed by NAO:
%		(A) Horizontal Normal arm movement (HN),
%		(B) Horizontal Faster arm movement (HF),
%		(C) Vertical Normal arm movement (VN) and
%		(D) Vertical Faster arm movement (VF).
%		Additionally,
%		(A) shows window sizes for 2-seconds (100 samples),
%		5-seconds (250 samples), 10-seconds (500 samples)
%		and 15-seconds (750 samples)
%		which are also presented in (B), (C) and (D).
%		Code and data to reproduce the figure is available in \cite{srep2021}.
%        }
%	\label{fig:sts}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------

\subsection*{Data collection from inertial measurement units} 
\label{sec:experiment:subsec:imu}
To give insight to the research questions, 
we considered various conditions of time series collected for this work: 
\begin{itemize}
\item Three levels of smoothness for normalised data 
	(e.g., sg0zmuv, sg1zmuv and sg2zmuv) where sg 
	stands for Savitzky-Golay filter with two different filter lengths (29 and 159) 
	and the same polynomial degree of 5 using the function \texttt{sgolay(p,n,m)} \cite{Rsignal}
	and zmuv is zero mean unit variance.
\item four arm movement activities with two velocities: 
	horizontal normal (HN), horizontal faster (HF), 
	vertical normal (VN), and vertical faster (VF), and
\item four window lengths: 2-sec (100 samples), 5-sec (250 samples), 
	10-sec (500 samples) and 15-sec (750 samples).
\end{itemize}
%After the data collection, raw time series were windowed, normalised and smoothed. 
Due to space limitations and to have simple visualisation, 
we only present 10-sec (500 samples) window length time series for 
three participants (p01, p01 and p03) performing horizontal 
arm movements (axis GyroZ) and vertical arm movements (axis GyroY) (Figs \ref{fig:ts}).
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/timeseries/pdf/fig1}
%    	\caption{
%	{\bf Time series for horizontal and vertical arm movements.}
%		(A/D) raw-normalised (sg0zmuv),
%		(B/E) normalised-smoothed 1 (sg1zmuv) and
%		(C/F) normalised-smoothed 2 (sg2zmuv).
%		Time series are only for three participants
%		($p01$, $p02$, and $p03$)
%		for horizontal and vertical arm movements in normal
%		and faster velocity (HN, HF, VN, VF)
%		with the normalised GyroZ or GyroY axis
%		and with one sensor attached to the participant (HS01)
%		and other sensor attached to the robot (RS01).
%	Code and data to reproduce the figure is available in \cite{srep2021}.
%       }
%    \label{fig:ts}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------

@@

\subsubsection*{Windowing data size}
With regards to the window size, Shoaib et al. in 2016 \cite{shoaib2016} 
investigated effects of seven window lengths (2, 5, 10, 15, 20, 25, 30 seconds)
and combination of inertial sensors (accelerometer, gyroscope and linear 
acceleration sensor) to improve the activity recognition performance for 
repetitive activities (walking, jogging and biking) and less repetitive 
activities (smoking, eating, giving a talk or drinking a coffee).
With that in mind, Shoaib et al. \cite{shoaib2016} concluded that the 
increase of window length improve the recognition of complex activities 
because these requires a large window length to learn the repetitive 
motion patterns. Particularly, one of the recommendations is to use large 
window size to recognise less repetitive activities which mainly involve 
random hand gestures. Therefore, for the four activities 
(HN, HF, VN, and VF) in this work, which are mainly repetitive, 
we select only four window sizes for analysis: 2-s window (100 samples), 
5-s window (250 samples), 10-s (500 samples) and 15-s window (750 samples) 
(Fig~\ref{fig:sts}).

@@

\subsubsection*{Uniform Time-Delay Embedding}
Using the overall embedding parameters ($\overline{m_0}=6$, $\overline{\tau_0}=8$), 
the first three axis of the rotated Principal Components Analysis (PCA) 
are shown for the reconstructed state spaces of horizontal (figures~\ref{fig:rsss}(A)) 
and vertical (figures~\ref{fig:rsss}(B)) arm movements. 
While visual inspection of these figures 
suggests differences in the trajectories of the reconstructed state spaces, 
we require an objective quantification to determine the extent of the differences.  
One approach could be use Euclidean distances from the origin for points in these 
trajectories, but this proved inconclusive and was not able to capture 
the dynamics of the trajectories. Consequently, we applied 
Recurrence Plots and Recurrence Quantification Analysis.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/versions/drawing-v00.pdf}
\caption{
	{\bf RSSs for horizontal and vertical arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig:ts}.
	Reconstructed state spaces were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$
	for (A) horizontal and (B) vertical arm movements.
	Code and data to reproduce the figure is available in \cite{srep2021}.	
        }
    \label{fig:rsss}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

>>>>

%*******************************************************************************
%*******************************************************************************
%*******************************************************************************
\section*{Experiment} \label{sec:experiment}
We conducted an experiment in the context of human-humanoid imitation (HHI) 
activities where participants were asked to imitate simple horizontal and 
vertical arm movements performed by NAO, a humanoid robot \cite{gouaillier2009}.
Such simple movements were repeated ten times for the participant 
who copied NAO's arm movements in a face-to-face imitation activity.
Also, wearable inertial measurement unit (IMU) sensors were attached 
to the right hand of the participant and to the left hand of the robot 
(Figure~\ref{fig02}A.1,C.1). Data were then collected with four NeMEMSi
IMU sensors with sampling rate of 50Hz provinding tri-axial data of the 
accelerometer, gyroscope and magnetometer sensors and quaternions 
\cite{Comotti2014}.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
  \centering
\includegraphics[width=1.0\textwidth]{figures/fig02/versions/drawing-v00.png}
    \caption{
	{\bf Human-humanoid imitation activities.} 
		Face-to-face human-humanoid imitation (HHI) activities for (A.1) HHI of horizontal arm movement, (B.1) Humanoid horizontal arm movement, (C.1) HHI of vertical arm movement, and (D.1) Humanoid vertical arm movement.
	{\bf Time series duration for arm movements.}
    	Time series of smoothed data from gyroscope sensor for different speed arm movements performed by NAO: (A.2) Horizontal Normal arm movement (HN), (B.2) Horizontal Faster arm movement (HF), (C.2) Vertical Normal arm movement (VN) and (D.2) Vertical Faster arm movement (VF).
		(A.2) shows window sizes for 2-seconds (100 samples), 5-seconds (250 samples), 10-seconds (500 samples) and 15-seconds (750 samples) which are also presented in (B.2), (C.2) and (D.2).
	{\bf Time series for horizontal and vertical arm movements.}
		(A.3/D.3) raw-normalised (sg0zmuv), (B.3/E.3) normalised-smoothed 1 (sg1zmuv) and (C.3/F.3) normalised-smoothed 2 (sg2zmuv).
		Time series are only for three participants	($p01$, $p02$, and $p03$) for horizontal and vertical arm movements in normal and faster velocity (HN, HF, VN, VF) with the normalised GyroZ or GyroY axis and with one sensor attached to the participant (HS01) and other sensor attached to the robot (RS01).
	Code and data to reproduce the figure is available in \cite{srep2021}.
}
    \label{fig02}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------

@@

\subsection*{Human-humanoid imitation activities}
For human-humanoid imitation (HHI) activities four neMEMSi sensors were used,
two of which were attached to the right hand of the participant and the 
other two to the left hand of the humanoid robot.
Then, each participant was asked to imitate repetitions of simple horizontal
and vertical arm movements performed by the humanoid robot in the following 
conditions:
(i) ten repetitions of horizontal arm movement at normal (HN) 
	and faster (HF) speed (Figure~\ref{fig02}A.1), and
(ii) ten repetitions of vertical arm movement at normal (VN) 
	and faster (VF) speed (Figure~\ref{fig02}C.1).
The normal and faster speed of arm movements is defined by the duration 
in number of samples of one repetition of NAO's arm movements. 
We select NAO's arm movements duration to distinguish between normal and 
faster arm movements as NAO's movements have less variation 
for such repetitive movements. 
The duration for one repetition of the horizontal 
arm movement at normal speed, HN, is 5 seconds considering 
that each repetition last around 250 samples.
For horizontal arm movement at faster speed, HF, each repetition were performed 
in 2 seconds which correspond to 90 samples of data.
The vertical arm movement at normal speed, VN, were performed  in 6 seconds 
which is around 300 samples of data. For vertical arm movement at 
faster speed, VF, each repetition lasts about 2.4 seconds which correspond 
to 120 samples of data.
To visualise the distinction between normal and faster speed for horizontal 
and vertical arm movements, Fig~\ref{fig02}A.2-D.2 shows smoothed time series
for axes Z and Y of the gyroscope sensors with four window lengths: 
2-sec (100-samples), 5-sec (250-samples), 10-sec (500-samples) 
and 15-sec (750-samples).
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/experiment/sts/pdf/sts}
%    \caption{
%	{\bf Time series duration of horizontal and vertical arm movements.}
%    		Time series of smoothed data from gyroscope sensor
%		for different speed arm movements performed by NAO:
%		(A) Horizontal Normal arm movement (HN),
%		(B) Horizontal Faster arm movement (HF),
%		(C) Vertical Normal arm movement (VN) and
%		(D) Vertical Faster arm movement (VF).
%		Additionally,
%		(A) shows window sizes for 2-seconds (100 samples),
%		5-seconds (250 samples), 10-seconds (500 samples)
%		and 15-seconds (750 samples)
%		which are also presented in (B), (C) and (D).
%		Code and data to reproduce the figure is available in \cite{srep2021}.
%        }
%	\label{fig:sts}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------

\subsection*{Data collection from inertial measurement units} 
\label{sec:experiment:subsec:imu}
To give insight to the research questions, 
we considered various conditions of time series collected for this work: 
\begin{itemize}
\item Three levels of smoothness for normalised data 
	(e.g., sg0zmuv, sg1zmuv and sg2zmuv) where sg 
	stands for Savitzky-Golay filter with two different filter lengths (29 and 159) 
	and the same polynomial degree of 5 using the function \texttt{sgolay(p,n,m)} \cite{Rsignal}
	and zmuv is zero mean unit variance.
\item four arm movement activities with two velocities: 
	horizontal normal (HN), horizontal faster (HF), 
	vertical normal (VN), and vertical faster (VF), and
\item four window lengths: 2-sec (100 samples), 5-sec (250 samples), 
	10-sec (500 samples) and 15-sec (750 samples).
\end{itemize}
%After the data collection, raw time series were windowed, normalised and smoothed. 
Due to space limitations and to have simple visualisation, 
we only present 10-sec (500 samples) window length time series for 
three participants (p01, p01 and p03) performing horizontal 
arm movements (axis GyroZ) and vertical arm movements (axis GyroY) (Figs \ref{fig02}A.3-F.3).
%%%---------------------------------(FIGURE)-------------------------------------
%\begin{figure}[ht]
%\centering
%\includegraphics[width=1.0\textwidth]{figures/timeseries/pdf/fig1}
%    	\caption{
%	{\bf Time series for horizontal and vertical arm movements.}
%		(A/D) raw-normalised (sg0zmuv),
%		(B/E) normalised-smoothed 1 (sg1zmuv) and
%		(C/F) normalised-smoothed 2 (sg2zmuv).
%		Time series are only for three participants
%		($p01$, $p02$, and $p03$)
%		for horizontal and vertical arm movements in normal
%		and faster velocity (HN, HF, VN, VF)
%		with the normalised GyroZ or GyroY axis
%		and with one sensor attached to the participant (HS01)
%		and other sensor attached to the robot (RS01).
%	Code and data to reproduce the figure is available in \cite{srep2021}.
%       }
%    \label{fig:ts}
%\end{figure}
%%%---------------------------------(FIGURE)------------------------------------

@@

\subsubsection*{Windowing data size}
With regards to the window size, Shoaib et al. in 2016 \cite{shoaib2016} 
investigated effects of seven window lengths (2, 5, 10, 15, 20, 25, 30 seconds)
and combination of inertial sensors (accelerometer, gyroscope and linear 
acceleration sensor) to improve the activity recognition performance for 
repetitive activities (walking, jogging and biking) and less repetitive 
activities (smoking, eating, giving a talk or drinking a coffee).
With that in mind, Shoaib et al. \cite{shoaib2016} concluded that the 
increase of window length improve the recognition of complex activities 
because these requires a large window length to learn the repetitive 
motion patterns. Particularly, one of the recommendations is to use large 
window size to recognise less repetitive activities which mainly involve 
random hand gestures. Therefore, for the four activities 
(HN, HF, VN, and VF) in this work, which are mainly repetitive, 
we select only four window sizes for analysis: 2-s window (100 samples), 
5-s window (250 samples), 10-s (500 samples) and 15-s window (750 samples) 
(Figures~\ref{fig02}A.2-D.2).

@@

\subsubsection*{Uniform Time-Delay Embedding}
Using the overall embedding parameters ($\overline{m_0}=6$, $\overline{\tau_0}=8$), 
the first three axis of the rotated Principal Components Analysis (PCA) 
are shown for the reconstructed state spaces of horizontal (figures~\ref{fig:rsss}(A)) 
and vertical (figures~\ref{fig:rsss}(B)) arm movements. 
While visual inspection of these figures 
suggests differences in the trajectories of the reconstructed state spaces, 
we require an objective quantification to determine the extent of the differences.  
One approach could be use Euclidean distances from the origin for points in these 
trajectories, but this proved inconclusive and was not able to capture 
the dynamics of the trajectories. Consequently, we applied 
Recurrence Plots and Recurrence Quantification Analysis.
%%---------------------------------(FIGURE)-------------------------------------
\begin{figure}[ht]
\centering
\includegraphics[width=1.0\textwidth]{figures/rss/versions/drawing-v00.pdf}
\caption{
	{\bf RSSs for horizontal and vertical arm movements.}
	Reconstructed state spaces for time series for p01 of Figure \ref{fig02}A.3-F.3.
	Reconstructed state spaces were computed with 
	embedding parameters 
	$\overline{m}_0=6$, $\overline{\tau}_0=8$
	for (A) horizontal and (B) vertical arm movements.
	Code and data to reproduce the figure is available in \cite{srep2021}.	
        }
    \label{fig:rsss}
\end{figure}
%%---------------------------------(FIGURE)------------------------------------
